Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-05-14 ms/spaceness.tex
demography lit updates

\begin{abstract}
    Real geography is continuous,
    but standard models in population genetics are based on discrete, well-mixed populations. 
    As a result many methods of analyzing genetic data assume that samples are a random draw from a well-mixed population, but are applied to clustered samples from populations that are structured clinally over space.
    Here we use simulations of populations living in continuous geography 
    to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and genome-wide association studies. 
    We find that several common summary statistics have distributions that differ substantially from that seen in well-mixed populations, 
    especially when Wright's neighborhood size is less than 100 and sampling is spatially clustered.
    The combination of low dispersal and clustered sampling causes demographic inference from the site frequency spectrum to infer more turbulent demographic histories, but averaged results across multiple simulations were surprisingly robust to isolation by distance. 
    We also show that the combination of spatially autocorrelated environments and limited dispersal
    can cause genome-wide association studies to identify numerous spurious signals of genetic association 
    with purely environmentally determined phenotypes,
    and that this bias is only partially corrected by regressing out principal components of ancestry. 
    Last, we discuss the relevance of our simulation results for inference from genetic variation in real organisms. 
\end{abstract}

@@

The assumption that populations are ``well-mixed'' has important implications for downstream inference of selection and demography. Methods based on the coalescent \citep{Kingman1982,wakeley2005coalescent} assume that the sampled individuals are a random draw from a well-mixed population
that is much larger than the sample \citep{wakeley2003gene}. The key assumption is that the individuals of each generation are \emph{exchangeable}, so that there is no correlation between the fate or fecundity of a parent and that of their offspring \citep{Huillet2011}. If dispersal or mate selection is limited by geographic proximity, this assumption can be violated in many ways. For instance, if mean viability or fecundity is spatially autocorrelated, then limited geographic dispersal will lead to parent--offspring correlations. Furthermore, nearby individuals will be more closely related than an average random pair, so drawing multiple samples from the same area of on the landscape will represent a biased sample of the genetic variation present in the whole population. It has long been appreciated that this model misspecification subjects downstream inferences to bias, but the extent and nature of these effects remain largely uninvestigated.

Two areas in which spatial structure may be particularly important are demographic inference and genome-wide association studies (GWAS). Effective population size is a fundamental parameter of population genetics and shapes expectations of the efficacy of selection and the rate of drift, among other evolutionary forces (\textit{c.f.} \cite{gillespie2001population}). Inferring a trajectory of population sizes through time has therefore become a major goal in population genetics, with many recent methods proposed to accomplish this task \citep{Liu2015,Terhorst2016,Schiffels2014,Sheehan2013}. All of these methods assume that the populations being modeled are approximately randomly mating (though this also depends on how researchers interpret the inferred parameter $\Theta=4N\mu$). Because isolation by distance is expected to change the rate of drift in a population \citep{Maruyama1972}, it may cause biases in inferred population sizes. However the effect of continuous spatial structure on demographic inference has not to our knowledge been previously investigated.

@@

Last, some of the most basic tools of population genetics are summary statistics like $F_{is}$ and Tajima's $D$, which are often interpreted as reflecting the influence of selection or demography on sampled populations \citep{tajima1989statistical}. Statistics like Tajima's $D$ are essentially summaries of the site frequency spectrum, which itself reflects variation in branch lengths and tree structure of the underlying genealogies of sampled individuals. Geographically limited mate choice distorts the distribution of these genealogies \citep{Maruyama1972}, which can affect the value of Tajima's $D$. Similarly, the distribution of tract lengths of identity by state among individuals contains information about not only historical demography \citep{Harris2013,ralph2013geography} and selection \citep{Garud2015}, but also dispersal and mate choice \citep{Ringbauer2017,Baharian2016}. 
We are particularly keen to examine how such summaries will be affected by models that incorporate continuous space, both to evaluate the assumptions underlying existing methods and to identify where the most promising signals of geography lie.

@@

Even though certain aspects of population demography depend on the scale of spatial interactions, it still could be that population genetic variation is well-described by a well-mixed population model. Indeed, mathematical results suggest that genetic variation in some spatial models should be well-approximated by a Wright-Fisher population if neighborhood size is large and all samples are geographically widely separated \citep{wilkins2004separation,zahle2005stepping}. However, the behavior of most common population genetic summary statistics has not yet been described in realistic geographic models. Moreover, as we will show, spatial sampling strategies can affect summaries of variation at least as strongly as the underlying population dynamics.

@@

Figure \ref{fig:sumstats} shows the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure~\ref{fig:sumstats}A) and several standard population genetic summary statistics (Figure~\ref{fig:sumstats}B). SFS comparisons show a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation, for smaller neighborhood sizes ($\leq 100$) that is exacerbated by midpoint and point sampling of individuals (depicted in Figure~\ref{fig:samplemap}). 
Reflecting this, Tajima's $D$ is quite positive in the same situations (Figure~\ref{fig:sumstats}B). Notably, the point at which Tajima's $D$ approaches 0 differs strongly across sampling strategies -- varying from a neighborhood size of roughly 50 for random sampling to at least 1000 for midpoint sampling.

One of the most commonly used summaries of variation is Tajima's summary of nucleotide divergence, $\theta_{\pi}$, calculated as the mean density of nucleotide differences, averaged across pairs of samples. As can be seen in Figure \ref{fig:sumstats}B, $\theta_{\pi}$ differs as much as nearly three-fold between the random mating and spatial models. $\theta_{\pi}$ using each sampling strategy approaches the random mating expectation at its own rate, but by a neighborhood size of around 100 all models are equivalent. The differences between spatial and random mating simulations are much greater than expected from differences in census size divided by variance in offspring number. This likely occurs because $\theta_{\pi}$ is a measure of mean time to most recent common ancestor between two samples, and at small values of $\sigma$, the time for dispersal to mix ancestry across the range exceeds the mean nonspatial coalescent time. (For instance, at the smallest value of $\sigma=0.2$, the range is 250 dispersal distances wide, and since the location of a diffusively moving lineage after $k$ generations has variance $k \sigma^2$, it takes around $250^2 = 62500$ generations to mix across the range, which is roughly ten times larger than the random mating effective population size.) Interestingly, the effect of sampling strategy is reversed relative to that observed in Tajima's D -- midpoint sampling reaches random mating expectations around neighborhood size 50, while random sampling is inflated until around neighborhood size 100.

@@

\subsection{Effects of Space on Demographic Inference}
One of the most important uses for population genetic data is inferring demographic history of populations. As demonstrated above, the site frequency spectrum and the distribution of IBS tracts varies across neighborhood sizes and sampling strategies. Does this variation lead to different inferences of past population sizes? To ask this we inferred population size histories from samples drawn from our simulated populations with two approaches: Stairwayplot \citep{Liu2015}, which uses a genome-wide estimate of the site frequency spectrum (SFS), and SMC++ \citep{Terhorst2016}, which incorporates information on both the SFS and linkage disequilibrium across the genome.

@@

However, individual model fits from both methods frequently reflected turbulent demographic histories (Figure \ref{fig:demography_supplement}), with the standard deviation of inferred $N_{e}$ across time points often exceeding the expected $N_{e}$ for both methods (Figure \ref{fig:demography}B). That is, despite the constant population sizes in our simulations, both methods tended to infer large fluctuations in population size over time, which could potentially result in incorrect biological interpretations. On average the variance of inferred population sizes was elevated at the lowest neighborhood sizes and declines as dispersal increases, with the strongest effects seen in stairwayplot model fits with for clustered sampling and neighborhood sizes less than 20 (Figure \ref{fig:demography}B). This suggests that very limited dispersal, particularly when combined with clustered sampling, may bias these methods towards inferring more variable demographic histories.

@@

\subsection{Effects of Sampling}
One of the most important differences between random mating and spatial models is the effect of sampling: 
in a randomly mating population the spatial distribution of sampling effort has no effect on estimates of genetic variation (Table \ref{table:sumstats}) , 
but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses. 
In most species,
the difficulty of traveling through all parts of a species range and the inefficiency of collecting single individuals at each sampling site 
means that most studies follow something closest to the ``point'' sampling strategy we simulated, 
in which multiple individuals are sampled from nearby points on the landscape. 
For example, in ornithology a sample of 10 individuals per species per locality is a common target when collecting for natural history museums. 
In classical studies of \textit{Drosophila} variation the situation is considerably worse, in which a single orchard might be extensively sampled.

@@

However, the largest sampling effects we observed occurred in our ``midpoint'' sampling strategy. 
This model is meant to reflect a bias in sampling effort towards the middle of a species' range. 
In empirical studies this sampling strategy could arise if, for example, 
researchers choose to sample the center of the range and avoid range edges to maximize probability of locating individuals during a short field season. 
Because midpoint sampling provides limited spatial resolution it dramatically reduces the magnitude of observed correlations between spatial and genetic distances. 
More surprisingly, midpoint sampling also leads to strongly positive Tajima's $D$ and an inflation in the proportion of heterozygous individuals in the sample. 
This increase in observed heterozygosity appears to reflect the effects of range edges, which are a fundamental facet of spatial genetic variation. 
If individuals move randomly in a finite two-dimensional landscape then regions in the middle of the landscape receive migrants from all directions while those on the edge receive no migrants from at least one direction. The average number of new mutations moving into the middle of the landscape is then higher than the number moving into regions near the range edge, leading to higher heterozygosity and lower inbreeding coefficients ($F_{IS}$) away from range edges. Though here we used only a single parameterization of fitness decline at range edges we believe this is a general property of non-infinite landscapes as it has also been observed in previous studies simulating under lattice models \citep{Neel2013,Shirk2014}.

@@

Classical results in population genetics collapse many aspects of life history into a single compound parameter, effective population size ($N_{e}$). Inferring effective population size back through time is now a common goal of population genomic analyses and an important step in establishing baseline expectations of genetic variation, e.g., when searching for signals of selection. Here we found that these methods are subject to bias when neighborhood sizes are less than 20 and sampling is spatially clustered, with Stairwayplot in particular inferring dramatic recent population bottlenecks for many simulations during the last 1000 generations. However we were surprised to observe that, at least when averaged across multiple simulations, the two methods we tested were relatively robust to violations of the assumption of random mating for most continuously structured populations with neighborhood sizes over 100. Indeed for SMC++ we found that the errors in the worst-affected bins had a roughly a two-fold underestimate of recent populations and a two-fold overestimate of ancient populations. Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms.

A more worrying pattern was the high level of variance in inferred $N_{e}$ trajectories for individual model fits using these methods, which was highest in simulations with the smallest neighborhood size. Often this reflected population crashes or booms in the recent past, but in practice it is rarely clear how far in the past we have to go before interpretting the $N_{e}$ values returned by these methods. This suggests that, at a minimum, researchers working with empirical data should replicate analyses multiple times and take a rolling average if model fits are inconsistent across runs. Splitting samples and running replicates on separate subsets -- the closest an empirical study can come to our design of averaging the results from multiple simulations -- may also alleviate this issue.

In general however, our analysis suggests that many empirical analyses of population size history using methods like SMC++ are somewhat robust to error caused by continuous spatial structure. Inferences drawn from static SFS-based methods like Stairwayplot should be treated with caution when there are signs of isolation by distance in the underlying data (for example, if a regression of $F_{st}$ against the logarithm of geographic distance has a significantly positive slope), and in particular an inference of population bottlenecks in the last 1000 years should be discounted if sampling is clustered, but estimates of deeper time patterns are likely to be fairly accurate. The biases in the SFS and haplotype structure we identified above are apparently small enough that they fall within the range of variability regularly inferred by these approaches, at least on datasets of the size we simulated.

@@

Over the last twenty years, genome-wide association studies (GWAS) have identified tens of thousands of correlations between genetic variation and phentoypes, both in humans and other species. Though originally intended to locate causal alleles for phenotypes controlled by a limited number of sites in the genome, GWAS is now routinely used as the first step in a phenotype-prediction pipeline that sums effect sizes across thousands or millions of variants to generate a "polygenic risk score" (PRS; \citep{Khera2018}). The most common method for estimating these effect sizes, which we followed in our analyses of simulated data here, is to regress the count of derived alleles at a site against individual phenotypes, taking the slope of this regression as an estimate of the effect of the allele on the phenotype. As recently reviewed by \cite{Visscher2019}, this is exactly the approach outlined by \cite{Fisher1918} at the dawn of quantitative genetics.

%Stepping back from the mechanics of GWAS specifically, Fisher's approach was to decompose the variance in phenotypes into environmental and genetic effects, e.g. 
%\begin{equation}
%P=G+E
%\end{equation}
%\begin{equation}
%var(P)=var(G)+var(E)+cov(G,E)
%\end{equation}
For GWAS in structured populations, much of the bias identified in previous studies \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015}, in which test statistics reflect population structure rather than phenotype association, arises because of the presence of a positive covariance between genotypic and environmental variation. When the environment and genotype covary across space their effects are confounded. Note this does not require interaction between genotype and environment (so-called GxE effects), which will introduce an additional source of bias. Here we refer simply to the fact that the allele frequency at a site will often covary with the environment when both breeding structure and the environment vary over space. To some degree the success of quantitative genetics in fields like agriculture may be due to the absence of this covariation -- crop and animal breeding operations can be conducted so that the environment is either controlled or measured across populations \citep{Wray2019}. However in natural populations this situation is extremely unlikely. Most populations are structured by a combination of limited dispersal and geographic barriers, and nearly all environments vary over space. GWAS in natural populations is then forced to confront the confounding effects of population structure and the environment directly.

Incorporating PC positions as covariates in the analysis \citep{Price2006} is designed to address this difficulty by regressing out a baseline level of ``average'' differentiation. 
In essence, a PC-corrected GWAS asks ``what regions of the genome are more associated with this phenotype than the average genome-wide association observed across populations?'' 
In our simulations, we observed that this procedure can fail under a variety of circumstances. 
If dispersal is limited and environmental variation is clustered in space (i.e., corner or patchy distributions in our simulations), PCA positions fail to capture the fine-scale spatial structure required to remove all signals of association. Conversely, as dispersal increases, PCA loses power to describe population structure before spatial mixing breaks down the relationship between genotype and the environment. These effects were observed with all spatially correlated environmental patterns, but were particularly pronounced if environmental effects are concentrated in one region, as was also found by \citet{Mathieson2012}.

Here we simulated a single chromosome with size roughly comparable to one human chromosome. If we scale the number of false-positive associations identified in our analyses to a GWAS conducted on whole-genome data from humans, we would expect to see several thousand weak false-positive associations after PC corrections in a population with neighborhood sizes up to at least 1000 (which should include values appropriate for many human populations). Notably, very few of the spurious associations we identified would be significant at a conservative Bonferroni-adjusted $p$-value cutoff (see Figure \ref{fig:manhattan_plots}). This suggests that GWAS focused on finding strongly associated alleles for traits controlled by a limited number of variants in the genome are likely robust to the impacts of continuous spatial structure. However, methods that analyze the combined effects of thousands or millions of weakly associated variants such as polygenic risk scores \citep{Khera2018} are likely to be affected by subtle population structure. Indeed as recently identified in studies of gentoype associations for human height in Europe \citep{Berg2018,Sohail2018}, PC regression GWAS in modern human populations do include residual signal of population structure in large-scale GWAS of polygenic traits. When attempting to make predictions across populations with different environmental exposures, PRS affected by population structure -- often termed "stratification bias" in the GWAS literature -- can be expected to offer low predictive power, as was shown in a recent study finding lower performance of PRS outside European populations \citep{Martin2019}.

A second point that has received less attention in the literature (but see \cite{Lawson2019}) is the issue of overfitting in GWAS. 
If a truly causal allele segregates at different frequencies in different populations, then correcting for population structure in a regression analysis will result in an underestimate of its effect size. 
Though our simulations had no causal alleles, we observed some evidence of this effect in the distribution of $p$-values across the genome (Figure \ref{fig:gwas}D): 
after PC regression many analyses resulted in $p$-values falling below their expected values from a uniform distribution. 
This is consistent with a recent empirical study of heritability in human height and body mass index, 
which found that increasing the number of PC axes used as covariates caused the total proportion of variance explained by SNPs to decline from $\approx0.8$ to $\approx0.75$ \citep{Wainschtein2019}.

>>>>

\begin{abstract}
    Real geography is continuous,
    but standard models in population genetics are based on discrete, well-mixed populations. 
    As a result many methods of analyzing genetic data assume that samples are a random draw from a well-mixed population, but are applied to clustered samples from populations that are structured clinally over space.
    Here we use simulations of populations living in continuous geography 
    to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and genome-wide association studies. 
    We find that most common summary statistics have distributions that differ substantially from that seen in well-mixed populations, 
    especially when Wright's neighborhood size is less than 100 and sampling is spatially clustered.
    The combination of low dispersal and clustered sampling causes demographic inference from the site frequency spectrum to infer more turbulent demographic histories, but averaged results across multiple simulations were surprisingly robust to isolation by distance. 
    We also show that the combination of spatially autocorrelated environments and limited dispersal
    causes genome-wide association studies to identify spurious signals of genetic association 
    with purely environmentally determined phenotypes,
    and that this bias is only partially corrected by regressing out principal components of ancestry. 
    Last, we discuss the relevance of our simulation results for inference from genetic variation in real organisms. 
\end{abstract}

@@

The assumption that populations are ``well-mixed'' has important implications for downstream inference of selection and demography. Methods based on the coalescent \citep{Kingman1982,wakeley2005coalescent} assume that the sampled individuals are a random draw from a well-mixed population
that is much larger than the sample \citep{wakeley2003gene}. The key assumption is that the individuals of each generation are \emph{exchangeable}, so that there is no correlation between the fate or fecundity of a parent and that of their offspring \citep{Huillet2011}. If dispersal or mate selection is limited by geographic proximity, this assumption can be violated in many ways. For instance, if mean viability or fecundity is spatially autocorrelated, then limited geographic dispersal will lead to parent--offspring correlations. Furthermore, nearby individuals will be more closely related than an average random pair, so drawing multiple samples from the same area of on the landscape will represent a biased sample of the genetic variation present in the whole population \citep{Stadler2009}.% It has long been appreciated that this model misspecification subjects downstream inferences to bias, but the extent and nature of these effects remain largely uninvestigated.

\textcolor{red}{Two areas in which spatial structure may be particularly important are demographic inference and genome-wide association studies (GWAS). Previous work has found that discrete population structure can create false signatures of population bottlenecks when attempting to infer demographic histories from microsatellite variation \citep{Chikhi2010}, statistics summarizing the site frequency spectrum (SFS) \citep{Ptak2002,Stadler2009,Onge2012}, or runs of homozygosity in a single individual \citep{Mazet2015}. The increasing availability of whole-genome data has led to the development of many methods that attempt to infer detailed trajectories of population sizes through time based on a variety of summaries of genetic data \citep{Liu2015,Schiffels2014,Sheehan2013,Terhorst2016}. 
Because all of these methods assume that the populations being modeled are approximately randomly mating, they are likely affected by spatial biases in the genealogy of sampled individuals \citep{Wakeley1999}, which may lead to incorrect inference of population changes over time \citep{Mazet2015}. However, previous investigations of these effects have focused on discrete rather than continuous space models, and the level of isolation by distance at which inference of population size trajectories become biased by structure is not well known. Here we test how two methods suitable for use with large samples of individuals -- Stairwayplot \citep{Liu2015} and SMC++ \citep{Terhorst2016} -- perform when applied to populations evolving in continuous space with varying sampling strategies and levels of dispersal. }

@@

Last, some of the most basic tools of population genetics are summary statistics like $F_{is}$ and Tajima's $D$, which are often interpreted as reflecting the influence of selection or demography on sampled populations \citep{tajima1989statistical}. Statistics like Tajima's $D$ are essentially summaries of the site frequency spectrum, which itself reflects variation in branch lengths and tree structure of the underlying genealogies of sampled individuals. Geographically limited mate choice distorts the distribution of these genealogies \citep{Maruyama1972,Wakeley1999}, which can affect the value of Tajima's $D$ \citep{Stadler2009}. Similarly, the distribution of tract lengths of identity by state among individuals contains information about not only historical demography \citep{Harris2013,ralph2013geography} and selection \citep{Garud2015}, but also dispersal and mate choice \citep{Ringbauer2017,Baharian2016}. 
We are particularly keen to examine how such summaries will be affected by models that incorporate continuous space, both to evaluate the assumptions underlying existing methods and to identify where the most promising signals of geography lie.

@@

Even though certain aspects of population demography depend on the scale of spatial interactions, it still could be that population genetic variation is well-described by a well-mixed population model. Indeed, mathematical results suggest that genetic variation in some spatial models should be well-approximated by a Wright-Fisher population if neighborhood size is large and all samples are geographically widely separated \citep{wilkins2004separation,zahle2005stepping}. However, the behavior of most common population genetic summary statistics \textcolor{red}{other than Tajima's $D$ \citep{Stadler2009}} has not yet been described in realistic geographic models. Moreover, as we will show, spatial sampling strategies can affect summaries of variation at least as strongly as the underlying population dynamics.

@@

Figure \ref{fig:sumstats} shows the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure~\ref{fig:sumstats}A) and several standard population genetic summary statistics (Figure~\ref{fig:sumstats}B). \textcolor{red}{Consistent with findings in n-island and stepping stone simulations \citep{Stadler2009}, the SFS shows a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation. This bias is most pronounced below neighborhood sizes $\leq 100$ and is exacerbated by midpoint and point sampling of individuals (depicted in Figure~\ref{fig:samplemap}).} 
Reflecting this, Tajima's $D$ is quite positive in the same situations (Figure~\ref{fig:sumstats}B). Notably, the point at which Tajima's $D$ approaches 0 differs strongly across sampling strategies -- varying from a neighborhood size of roughly 50 for random sampling to at least 1000 for midpoint sampling.

One of the most commonly used summaries of variation is Tajima's summary of nucleotide divergence, $\theta_{\pi}$, calculated as the mean density of nucleotide differences averaged across pairs of samples. As can be seen in Figure \ref{fig:sumstats}B, $\theta_{\pi}$ differs as much as nearly three-fold between the random mating and spatial models. $\theta_{\pi}$ using each sampling strategy approaches the random mating expectation at its own rate, but by a neighborhood size of around 100 all models are equivalent. The differences between spatial and random mating simulations are much greater than expected from differences in census size divided by variance in offspring number. This likely occurs because $\theta_{\pi}$ is a measure of mean time to most recent common ancestor between two samples, and at small values of $\sigma$, the time for dispersal to mix ancestry across the range exceeds the mean nonspatial coalescent time. (For instance, at the smallest value of $\sigma=0.2$, the range is 250 dispersal distances wide, and since the location of a diffusively moving lineage after $k$ generations has variance $k \sigma^2$, it takes around $250^2 = 62500$ generations to mix across the range, which is roughly ten times larger than the random mating effective population size.) Interestingly, the effect of sampling strategy is reversed relative to that observed in Tajima's D -- midpoint sampling reaches random mating expectations around neighborhood size 50, while random sampling is inflated until around neighborhood size 100.

@@

\subsection{Effects of Space on Demographic Inference}
One of the most important uses for population genetic data is inferring demographic history of populations. As demonstrated above, the site frequency spectrum and the distribution of IBS tracts varies across neighborhood sizes and sampling strategies. Does this variation lead to different inferences of past population sizes? To ask this we inferred population size histories from samples drawn from our simulated populations with two approaches: Stairwayplot \citep{Liu2015}, which uses a genome-wide estimate of the SFS, and SMC++ \citep{Terhorst2016}, which incorporates information on both the SFS and linkage disequilibrium across the genome.

@@

However, individual model fits from both methods frequently reflected turbulent demographic histories (Figure \ref{fig:demography_supplement}), with the standard deviation of inferred $N_{e}$ across time points often exceeding the expected $N_{e}$ for both methods (Figure \ref{fig:demography}B). That is, despite the constant population sizes in our simulations, both methods tended to infer large fluctuations in population size over time, which could potentially result in incorrect biological interpretations. On average the variance of inferred population sizes was elevated at the lowest neighborhood sizes and declines as dispersal increases, with the strongest effects seen in stairwayplot model fits with for clustered sampling and neighborhood sizes less than 20 (Figure \ref{fig:demography}B).

@@

\subsection{Effects of Sampling}

One of the most important differences between random mating and spatial models is the effect of sampling: 
in a randomly mating population the spatial distribution of sampling effort has no effect on estimates of genetic variation (Table \ref{table:sumstats}) , 
but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses \textcolor{red}{(see also \cite{Stadler2009})}. 
In most species,
the difficulty of traveling through all parts of a species range and the inefficiency of collecting single individuals at each sampling site 
means that most studies follow something closest to the ``point'' sampling strategy we simulated, 
in which multiple individuals are sampled from nearby points on the landscape. 
For example, in ornithology a sample of 10 individuals per species per locality is a common target when collecting for natural history museums. 
In classical studies of \textit{Drosophila} variation the situation is considerably worse, in which a single orchard might be extensively sampled.

@@

However, the largest sampling effects we observed occurred in our ``midpoint'' sampling strategy. 
This model is meant to reflect a bias in sampling effort towards the middle of a species' range. 
In empirical studies this sampling strategy could arise if, for example, 
researchers choose to sample the center of the range and avoid range edges to maximize probability of locating individuals during a short field season. 
Because midpoint sampling provides limited spatial resolution it dramatically reduces the magnitude of observed correlations between spatial and genetic distances. 
More surprisingly, midpoint sampling also leads to strongly positive Tajima's $D$ and an inflation in the proportion of heterozygous individuals in the sample -- similar to the effect of sampling a single deme in an n-island model as reported in \citep{Stadler2009}. 
This increase in observed heterozygosity appears to reflect the effects of range edges, which are a fundamental facet of spatial genetic variation. 
If individuals move randomly in a finite two-dimensional landscape then regions in the middle of the landscape receive migrants from all directions while those on the edge receive no migrants from at least one direction. The average number of new mutations moving into the middle of the landscape is then higher than the number moving into regions near the range edge, leading to higher heterozygosity and lower inbreeding coefficients ($F_{IS}$) away from range edges. Though here we used only a single parameterization of fitness decline at range edges we believe this is a general property of non-infinite landscapes as it has also been observed in previous studies simulating under lattice models \citep{Neel2013,Shirk2014}.

@@

\textcolor{red}{Previous studies have found that population structure and nonrandom sampling can create spurious signals of population bottlenecks when attempting to infer demographic history with microsatellite variation, summary statistics, or runs of homozygosity \citep{Chikhi2010,Stadler2009,Ptak2002,Mazet2015}. Here we found that methods that infer detailed population trajectories through time based on the SFS and patterns of LD across the genome are also subject to this bias, with some combinations of dispersal and sampling strategy inferring deep recent population bottlenecks and overestimating ancient $N_{e}$ by a around a factor of 2. However we were surprised to see that both Stairwayplot and SMC++ can tolerate relatively strong isolation by distance -- i.e. neighborhood sizes of 20 -- and still perform well when averaging results across multiple simulations. Inference in populations with neighborhood sizes over 20 was relatively accurate unless samples were concentrated in the middle of the range (Figure \ref{fig:demography}). As we will discuss below, most empirical estimates of neighborhood size, including all estimates for human populations, are large enough that population size trajectories inferred by these approaches should not be strongly affected by spatial biases created by dispersal in continuous landscapes. In contrast, \cite{Mazet2015} found that varying migration rates through time could create strong biases in inferred population trajectories from an $n$-island model with parameters relevant for human history, suggesting that changes in migration rates through time are more likely to drive variation in inferred $N_{e}$ than isolation by distance.}

\textcolor{red}{Between the two methods we tested, we found that SMC++ was more robust to spatial biases than Stairwayplot, with the worst-affected bins underestimating recent populations by roughly half rather than nearly 10-fold as with Stairwayplot.} Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms. A more worrying pattern was the high level of variance in inferred $N_{e}$ trajectories for individual model fits using these methods, which was highest in simulations with the smallest neighborhood size (Figure \ref{fig:demography}, Figure \ref{fig:demography_supplement}). This suggests that, at a minimum, researchers working with empirical data should replicate analyses multiple times and take a rolling average if model fits are inconsistent across runs. Splitting samples and running replicates on separate subsets -- the closest an empirical study can come to our design of averaging the results from multiple simulations -- may also alleviate this issue.

In general however, our analysis suggests that many empirical analyses of population size history using methods like SMC++ are robust to error caused by spatial structure within continuous landscapes. Inferences drawn from static SFS-based methods like Stairwayplot should be treated with caution when there are signs of isolation by distance in the underlying data (for example, if a regression of $F_{st}$ against the logarithm of geographic distance has a significantly positive slope), and in particular an inference of population bottlenecks in the last 1000 years should be discounted if sampling is clustered, but estimates of deeper time patterns are likely to be fairly accurate. The biases in the SFS and haplotype structure identified above (see also \citep{Wakeley1999,Chikhi2010,Stadler2009}) are apparently small enough that they fall within the range of variability regularly inferred by these approaches, at least on datasets of the size we simulated.

@@

\textcolor{red}{Spatial structure is particularly challenging for genome-wide association studies, because the effects of dispersal on genetic variation are compounded by spatial variation in the environment \citep{Mathieson2012}. Spatially restricted mate choice and dispersal causes variation in allele frequencies across the range of a species. If environmental factors affecting the phenotype of interest also vary over space, then groups of individuals in different regions will both experience different environments and have different allele frequencies at many sites in the genome. In this scenario an uncorrected GWAS will infer genetic associations with a purely environmental phenotype at any site in the genome that is differentiated over space, and the relative degree of bias will be a function of the degree of covariation in allele frequencies and the environment (i.e. Figure \ref{fig:gwas}C, bottom panel). This pattern is now well known and has been demonstrated in a variety of simulation and empirical contexts \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015,Berg2018,Sohail2018}.}

Incorporating PC positions as covariates in a linear-regression GWAS \citep{Price2006} is designed to address this challenge by regressing out a baseline level of ``average'' differentiation. 
In essence, a PC-corrected GWAS asks ``what regions of the genome are more associated with this phenotype than the average genome-wide association observed across populations?'' 
In our simulations, we observed that this procedure can fail under a variety of circumstances. 
If dispersal is limited and environmental variation is clustered in space (i.e., corner or patchy distributions in our simulations), PCA positions fail to capture the fine-scale spatial structure required to remove all signals of association. Conversely, as dispersal increases, PCA loses power to describe population structure before spatial mixing breaks down the relationship between genotype and the environment. These effects were observed with all spatially correlated environmental patterns, but were particularly pronounced if environmental effects are concentrated in one region, as was also found by \citet{Mathieson2012}. Though increasing the number of PC axes used in the analysis may reduce the false-positive rate, this may also decrease the power of the test to detect truly causal alleles \citep{Lawson2019}.

Here we simulated a single chromosome with size roughly comparable to one human chromosome. If we scale the number of false-positive associations identified in our analyses to a GWAS conducted on whole-genome data from humans, we would expect to see several thousand weak false-positive associations after PC corrections in a population with neighborhood sizes up to at least 1000 (which should include values appropriate for many human populations). Notably, very few of the spurious associations we identified would be significant at a conservative Bonferroni-adjusted $p$-value cutoff (see Figure \ref{fig:manhattan_plots}). This suggests that GWAS focused on finding strongly associated alleles for traits controlled by a limited number of variants in the genome are likely robust to the impacts of continuous spatial structure. However, methods that analyze the combined effects of thousands or millions of weakly associated variants such as polygenic risk scores \citep{Khera2018} are likely to be affected by subtle population structure. Indeed as recently identified in studies of gentoype associations for human height in Europe \citep{Berg2018,Sohail2018}, PC regression GWAS in modern human populations do include residual signal of population structure in large-scale analyses of polygenic traits. When attempting to make predictions across populations with different environmental exposures, PRS affected by population structure can be expected to offer low predictive power, as was shown in a recent study finding lower performance of PRS outside European populations \citep{Martin2019}.

In summary, spatial covariation in population structure and the environment confound the interpretation of GWAS $p$-values, and correction using principal components is insufficient to fully separate these signals for polygenic traits under a variety of environmental and population parameter regimes. 
Other GWAS methods may be less sensitive to this confounding, but there is no obvious reason that this should be so.
One approach to estimating the degree of bias in GWAS caused by population structure is LD score regression \citep{Bulik-Sullivan2015}. 
Though this approach appears to work well in practice, its interpretation is not always straightforward and it is likely biased by the presence of linked selection \citep{Berg2018}. \textcolor{red}{In addition, we observed that in many cases the false-positive SNPs we identified appeared to be concentrated in LD peaks similar to those expected from truly causal sites (Figure \ref{fig:manhattan_plots}), which may confound LD score regression.}
We suggest a straightforward alternative for species in which the primary axes of population differentiation is space 
(note this is likely not the case for some modern human populations): 
run a GWAS with spatial coordinates as phenotypes and check for $p$-value inflation or significant associations. 
If significant associations with sample locality are observed after correcting for population structure, 
the method is sensitive to false positives induced by spatial structure. 
This is essentially the approach taken in our ``clinal'' model (though we add normally distributed noise to our phenotypes). 
Of course, it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.

%%%%%% old text from gwas section
%In a classic quantitative genetics framework phenotypes are the sum of genetic and environmental effects, $P=G+E$, with variance $var(P)=var(G)+var(E)+cov(G,E)$. Much of the bias identified in previous studies \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015}, in which GWAS test statistics reflect population structure rather than phenotype association, arises because of a positive covariance between genotypic and environmental variation (the third term above). Because the environment is typically not measured in GWAS of natural populations, this covariation confounds genetic and environmental effects on the phenotype of interest and leads to overestimated effect sizes and deflated $p$ values.

%Over the last twenty years, genome-wide association studies (GWAS) have identified tens of thousands of correlations between genetic variation and phentoypes, both in humans and other species. Though originally intended to locate causal alleles for phenotypes controlled by a limited number of sites in the genome, GWAS is now routinely used as the first step in a phenotype-prediction pipeline that sums effect sizes across thousands or millions of variants to generate a "polygenic risk score" (PRS; \citep{Khera2018}). The most common method for estimating these effect sizes, which we followed in our analyses of simulated data here, is to regress the count of derived alleles at a site against individual phenotypes, taking the slope of this regression as an estimate of the effect of the allele on the phenotype. As recently reviewed by \cite{Visscher2019}, this is exactly the approach outlined by \cite{Fisher1918} at the dawn of quantitative genetics.
